Digital enhancement of the ``Paolo Orsi'' museum: a Google Street View 360° pilot project tour

### 2016-01-19 source/13/EAGLE2016_Full_Paper_Verreth.tex
papers revised
minor fixes

 The main credits for the whole set-up of this new project go to Mark Depauw, the Trismegistos director. The general
idea remained the tagging of all words starting with a capital, but this time Depauw tried a new approach, which was
completely imbedded within Filemaker. Since the Roman 'tria nomina' hardly occur in the Greek papyri, also a new
onomastical structure had to be devised for the automatic recognition. As a test case we choose the full text corpus of
the EDCS. All these texts were 'cut up' in capital clusters, i.e. strings of consecutive words all starting with a
capital. Words as filius, nepos and libertus were added to the string so that in most cases the full identification of
a person could be grouped together, e.g. Quintus Caecilius Quinti filius Quirina Mustacus (TM 332260), Caio Annioleno
Cai filio Arnensi Karthaginiensi Galliano (TM 349961) or Maesiaes Cai libertae Chrysidis (TM 244384). A minor
disadvantage of this corpus was the capitalisation of the first word of every inscription (which is not done in the
PN), which yields quite a significant number of mere nouns in the group of expected personal names and toponyms. Also
other words starting with a capital were not our prime goal: names of gods, religious festivals, months, army units,
ships, animals, mythological persons and Roman numbers. Excluded (for the moment) are also the names of the emperors
and the members of the imperial family, often occurring in a complicated titulature which is not easy to standardise.
In the end this yielded 898.134 capital cluster cards. From our previous projects we already had a fairly elaborated
reference corpus of Roman personal names, which was now expanded and used by Depauw to match every word in the capital
clusters with the names in the reference corpus. If there was a match, the case of the ending was added, e.g.

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img001.png}
\caption{default}
\label{fig:1}
\end{figure}

@@

This is also the phase where the toponyms come in (which can occur both in the first and in the second group).
Within the EAGLE project we created already a fairly large reference corpus for toponyms from the Roman empire, but
this corpus was enlarged by entering the toponyms occurring in the Itinerarium provinciarum Antonini Augusti [3.434]
and the Tabula Peutingeriana [3.287]. The TM Geo file now contains 46.277 toponyms from all over Egypt and the Roman
empire, both ancient and modern, which cover most of the places where ancient texts have been found, and most of the
toponyms mentioned in Egyptian papyrological and Latin epigraphical sources.\footnote{ For more information on TM Geo,
\citep{verreth2013}.} We played with the idea of automatic matching, like we did for the personal
names, but except for the case of the relatively straightforward tribus names, this yielded no satisfactory results. A
lot of toponyms resemble personal names (e.g. Florentia, Venusia, (Fundus) Bassianus), and the automated identification
of strings such as Colonia Ulpia Traiana Augusta Fidelis Lepcis Magna (TM 198383) or Municipium Augustum Hipponiensium
Regiorum (TM 200133) seemed too cumbersome. In the end, we settled for plan B, which shows that in `Digital
Humanities', the human component is still essential: we had to go through all 900.000 capital cluster cards manually,
identifying the toponyms in every cluster, adding the corresponding TM Geo number and - whenever necessary - correcting
the indications for yes/no and the automatic identification of the cases of the personal names. It was six months of
tedious work, resulting in 61.139 capital clusters with at least one toponym. No doubt some toponyms escaped our
attention, but we do hope to have identified the majority of the names involved. In this process, however, we also
encountered some set backs, especially in the longer texts and in the more complicated wooden or metal tablets: in the
beta version on which we worked, the creation of the capital cluster strings was not always so perfect as we had hoped
for, and also the line numbers automatically assigned to each cluster string have sometimes gone astray. Mark Depauw is
developing a new and improved version, especially in preparation for the much larger batch of personal names, where it
is virtually impossible to manually correct everything that has gone wrong. Due to these problems I guess that we now
have about only 80 to 90 \% of the toponyms occurring in all Latin inscriptions, but on the whole we are quite pleased
with the result and in due time the remaining toponyms no doubt will find their way into the database also.

Phase 1, the identification of toponyms in the capital cluster strings, was finished in the beginning of July 2015,
and we have now entered phase 2, the incorporation of the capital cluster file into the `real' TM Georef file. For
every place listed in TM Geo we try to list all the ancient text references where that place is attested. The file of
these geographical references (TM Georef) is directly linked with the main TM Text file, so that every reference
automatically receives a chronological and a geographical context. Every toponym found in the capital cluster file is
exported to a separate Georef card. When a toponym exists out of several consecutive elements, like the colonia and
municipium examples mentioned before, they are automatically grouped on one card. Twofold toponyms such as `Bythina et
Ponthus', which cannot belong to the same capital cluster string because of the intermediary `et', are exported double
and then afterwards joined manually. For the moment TM Georef lists 67.820 geographical references attested in Latin
documentary texts and 10.474 references attested in Latin literary texts (but except for Egypt the latter have not yet
systematically been entered). The Latin references make out almost 40 \% of the total of 196.794 Georef cards.

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img002.png}
\caption{default}
\label{fig:2}
\end{figure}

@@

Since the users of TM must be able to look for specific spelling variants of each toponyms, all these references are
presented the way they are on the stone, with as little additions or emendations as possible, except of course for any
abbreviations at the end of the word; e.g. T(h)ra[c(um)] becomes Tra[c(um)], and Rom(a)e becomes Rome. In another field
the standardized nominative case is given, without brackets or uncertainty dots; e.g. Trax and Roma.

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img003.png}
\caption{default}
\label{fig:3}
\end{figure}

@@

The third and final phase involves the context of the toponym. In the field Detail we give a plain translation of
the immediate phrase to which the toponym belongs. The translation should be a standardised as possible, with termini
technici preferably added in Latin, so that the users can easily search for them in the database; e.g. `Tiberius Iulius
Martialis son of Tiberius of (the tribus) Claudia from Savaria, soldier (miles) of legio XV Apollinaris', `praefectus
Aegypti', `cohors I Tungrorum'. If the place is explicitly ascribed to a provincia or a region, this provincia is
listed in the field 'Administrative situation'. If a town is explicitly called an oppidum, a vicus or a civitas, this
information is listed in the field Status. By adding this information in searchable fields we hope that the user can
start asking quite specific questions; e.g. the first and last attestation of a place in the sources; the periods in
which a town was called a colonia or a municipium; the places in which the `ala I Thracum Mauretana' has been attested.

>>>>

 The main credits for the whole set-up of this new project go to Mark Depauw, the Trismegistos director. The general
idea remained the tagging of all words starting with a capital, but this time Depauw tried a new approach, which was
completely imbedded within Filemaker. Since the Roman 'tria nomina' hardly occur in the Greek papyri, also a new
onomastical structure had to be devised for the automatic recognition. As a test case we choose the full text corpus of
the EDCS. All these texts were 'cut up' in capital clusters, i.e. strings of consecutive words all starting with a
capital. Words as \emph{filius}, \emph{nepos} and \emph{libertus} were added to the string so that in most cases the full identification of
a person could be grouped together, e.g. \emph{Quintus Caecilius Quinti filius Quirina Mustacus} (TM 332260), \emph{Caio Annioleno
Cai filio Arnensi Karthaginiensi Galliano} (TM 349961) or \emph{Maesiaes Cai libertae Chrysidis} (TM 244384). A minor
disadvantage of this corpus was the capitalization of the first word of every inscription (which is not done in the
PN), which yields quite a significant number of mere nouns in the group of expected personal names and toponyms. Also
other words starting with a capital were not our prime goal: names of gods, religious festivals, months, army units,
ships, animals, mythological persons and Roman numbers. Excluded (for the moment) are also the names of the emperors
and the members of the imperial family, often occurring in a complicated titulature which is not easy to standardise.
In the end this yielded 898.134 capital cluster cards. From our previous projects we already had a fairly elaborated
reference corpus of Roman personal names, which was now expanded and used by Depauw to match every word in the capital
clusters with the names in the reference corpus. If there was a match, the case of the ending was added, e.g.

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img001.png}
\caption{Named Entity Recognition with FileMaker Pro 14}
\label{fig:1}
\end{figure}

@@

This is also the phase where the toponyms come in (which can occur both in the first and in the second group).
Within the EAGLE project we created already a fairly large reference corpus for toponyms from the Roman empire, but
this corpus was enlarged by entering the toponyms occurring in the \emph{Itinerarium provinciarum Antonini Augusti} [3.434]
and the \emph{Tabula Peutingeriana} [3.287]. The TM Geo file now contains 46.277 toponyms from all over Egypt and the Roman
empire, both ancient and modern, which cover most of the places where ancient texts have been found, and most of the
toponyms mentioned in Egyptian papyrological and Latin epigraphical sources.\footnote{ For more information on TM Geo,
\citep{verreth2013}.} We played with the idea of automatic matching, like we did for the personal
names, but except for the case of the relatively straightforward tribus names, this yielded no satisfactory results. A
lot of toponyms resemble personal names (e.g. \emph{Florentia}, \emph{Venusia}, \emph{(Fundus) Bassianus}), and the automated identification
of strings such as \emph{Colonia Ulpia Traiana Augusta Fidelis Lepcis Magna }(TM 198383) or \emph{Municipium Augustum Hipponiensium
Regiorum} (TM 200133) seemed too cumbersome. In the end, we settled for plan B, which shows that in `Digital
Humanities', the human component is still essential: we had to go through all 900.000 capital cluster cards manually,
identifying the toponyms in every cluster, adding the corresponding TM Geo number and - whenever necessary - correcting
the indications for yes/no and the automatic identification of the cases of the personal names. It was six months of
tedious work, resulting in 61.139 capital clusters with at least one toponym. No doubt some toponyms escaped our
attention, but we do hope to have identified the majority of the names involved. In this process, however, we also
encountered some set backs, especially in the longer texts and in the more complicated wooden or metal tablets: in the
beta version on which we worked, the creation of the capital cluster strings was not always so perfect as we had hoped
for, and also the line numbers automatically assigned to each cluster string have sometimes gone astray. Mark Depauw is
developing a new and improved version, especially in preparation for the much larger batch of personal names, where it
is virtually impossible to manually correct everything that has gone wrong. Due to these problems I guess that we now
have about only 80 to 90 \% of the toponyms occurring in all Latin inscriptions, but on the whole we are quite pleased
with the result and in due time the remaining toponyms no doubt will find their way into the database also.

Phase 1, the identification of toponyms in the capital cluster strings, was finished in the beginning of July 2015,
and we have now entered phase 2, the incorporation of the capital cluster file into the `real' TM Georef file. For
every place listed in TM Geo we try to list all the ancient text references where that place is attested. The file of
these geographical references (TM Georef) is directly linked with the main TM Text file, so that every reference
automatically receives a chronological and a geographical context. Every toponym found in the capital cluster file is
exported to a separate Georef card. When a toponym exists out of several consecutive elements, like the \emph{colonia} and
\emph{municipium} examples mentioned before, they are automatically grouped on one card. Twofold toponyms such as `\emph{Bythina et
Ponthus'}, which cannot belong to the same capital cluster string because of the intermediary `\emph{et}', are exported double
and then afterwards joined manually. For the moment TM Georef lists 67.820 geographical references attested in Latin
documentary texts and 10.474 references attested in Latin literary texts (but except for Egypt the latter have not yet
systematically been entered). The Latin references make out almost 40 \% of the total of 196.794 Georef cards.

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img002.png}
\caption{The Georef Card}
\label{fig:2}
\end{figure}

@@

Since the users of TM must be able to look for specific spelling variants of each toponyms, all these references are
presented the way they are on the stone, with as little additions or emendations as possible, except of course for any
abbreviations at the end of the word; e.g. \emph{T(h)ra[c(um)]} becomes \emph{Tra[c(um)]}, and Rom(a)e becomes \emph{Rome}. In another field the standardized nominative case is given, without brackets or uncertainty dots; e.g. \emph{Trax} and \emph{Roma}.

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img003.png}
\caption{Philadelphia}
\label{fig:3}
\end{figure}

@@

The third and final phase involves the context of the toponym. In the field Detail we give a plain translation of
the immediate phrase to which the toponym belongs. The translation should be a standardised as possible, with termini
technici preferably added in Latin, so that the users can easily search for them in the database; e.g. `\emph{Tiberius Iulius
Martialis son of Tiberius of (the tribus) Claudia from Savaria, soldier (miles) of legio XV Apollinaris}', `\emph{praefectus
Aegypti}', `\emph{cohors I Tungrorum}'. If the place is explicitly ascribed to a \emph{provincia} or a region, this \emph{provincia} is
listed in the field 'Administrative situation'. If a town is explicitly called an \emph{oppidum}, a \emph{vicus} or a \emph{civitas}, this
information is listed in the field Status. By adding this information in searchable fields we hope that the user can
start asking quite specific questions; e.g. the first and last attestation of a place in the sources; the periods in
which a town was called a \emph{colonia} or a \emph{municipium}; the places in which the `\emph{ala I Thracum Mauretana}' has been attested.

### 2016-01-19 source/20/EAGLE2016_BONACINI - Pilot project at Paolo Orsi Museum.tex
papers revised
minor fixes

\title{Digital enhancement of the ``Paolo Orsi'' museum: a Google Street View 360° pilot project tour}
\author[catania]{Elisa Bonacini\corref{first}}
\cortext[first]{Corresponding author. Email: e_bonacini@hotmail.com; elisa.bonacini@unict.it}
\address[catania]{Humanities Department, University of Catania; IEMEST (Istituto Euro Mediterraneo di Scienze e Tecnologie), Palermo, Italy}
\date{2015-12-11}
\begin{document}

@@

Sicily has the highest number of UNESCO heritage sites (7/51 in total\footnote{1997: Valley of Temples in Agrigento;
1997: Villa del Casale; 2000: Eolian Islands; 2002: Late Baroque Towns of the Val di Noto (South-Eastern Sicily); 2005:
Syracuse and the rock necropolisof Pantalica; 2013: Mount Etna; 2015: Arab-Norman sites, Palermo and the Cathedral
Churches of Cefalù and Monreale.}) and of UNESCO intangible cultural heritage (3/6\footnote{2008: Opera dei Pupi,
Sicilian puppet theatre; 2013: Mediterranean diet (transnational); 2014: Traditional agricultural practice of
cultivating the `vite ad alberello' (head-trained bush vines) of the community of Pantelleria.}) in Italy and in the
world. Infact, Sicily has some of the largest and most important archaeological sites in the world: the temple of
Concordia in Agrigento, for its exceptional state of preservation has become the symbol of UNESCO itself.

Despite of this, Sicilian cultural heritage struggles to be present on Google’s platforms such as Google Art Project and
others, as it should, compared to other Italian cultural sites \citep{Bonacini2013} \citep{Bonacini2014}).

@@

In Street View Gallery\footnote{\url{https://www.google.com/maps/views/streetview?gl=us}.}, which now has contributed a great
number of users from all over the world, thousands are the spherical and geo located photos of Italian places. However,
tightening the selection to ``Landmarks of Italy''
\footnote{\url{https://www.google.com/maps/views/streetview/italy-highlights?gl=us}.}, Sicily has only 9 spherical photos,
showing the beaches of the Aeolian Islands (7), the islands of Favignana (1) and Marettimo (1).

Among Art Project’s 605 museum collections\footnote{\url{https://www.google.com/maps/views/streetview/art-project?gl=it}.}, 47
are Italian, especially relevant to Rome, Turin, Venice and Milan. One of these, not the best artistic production, is a
Sicilian contemporary collection, relevant to the International Festival of Street Artists in Giardini Naxos (Me). No
other Sicilian museum, collection or archaeological site has been included in Art Project.

Browsing on the Street View Gallery, 21 sites are all over inscribed on the World Wonders
Project\footnote{\url{https://www.google.com/maps/views/streetview/world-wonders-project?gl=it}.}: for Italy, only Pompeii
and the historic center of Florence are inscribed in this still restricted list. Many more sites can be visited
virtually from the same project linked on Google Cultural Institute’s
website\footnote{\url{www.google.com/intl/it/culturalinstitute/worldwonders/}. }: 172 sites in the world, 22 in Italy and,
finally, 2 of them in Sicily. Until last year the Unesco site of the baroque town Val di Noto was the only one in
Sicily; currently Mount Etna, has been added.

@@

Art Project, for example, allows you to view points of interest along the path in a museum, but captions are short and
photos are static. A pilot initiative has been recently launched featuring the possibility to in-browse almost 300 3D
photos of objects from the collections of six cultural institutions in the world. However 242 of these consist of scans
of animal skulls (pieces of nature, not works of art) from the California Academy of Science collection; 22 objects of
art come from Museo d’Arte Orientale (the only Italian cultural institution to join the project until now). These 3D
objects, described by short captions, are available to users to be rotated and zoomed. The aim of this new Google’s
project is to begin building the most important and largest database of 3D scan art of works worldwide.

@@

As for the displaying on Google Maps of the 360° virtual tour of the objects, we must specify that Maps so far does not
support the integration of menus, captions, photos, video, info inside the Street View virtual tour technology.
Infact,it is only allowed navigation in 360°, ie a virtual walk.

@@

The B2 Sector introduces the visitor to the archaeological finds from the city of Syracuse, from its foundation to
classical age. Here the most important spaces are those dedicated to the architecture of archaic and classical temples
(Athenaion, Ionic temple, Olympeion and Apollonion), to the statuary and the terracotta findings from urban excavations
during the last decades (Piazza Duomo, Ortigia, Piazza della Vittoria) and to the urban and ex-traurban necropolis
(Fusco, Giardino Spagna) with rich funerary kits. \

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img004.jpg}
\caption{360° tour of Sector D: the Venus Landolina.}
\label{fig:}
\end{figure}

@@

\begin{otherlanguage*}{greek}
chrismon ς χριστ<ι>ανῆς σεμνῆς ἀγανóφρονος [ἠδ]ὲ φιλάνδρου Nασσιανῆς τύμβον εἰσορᾷς, φίλε, κίμενον [ὣδε ἥ]τις σεμνοσύνῃσιν [ἐριζε]το Πηνελοπίῃ palma chrismon ἐνθάδε κῖτε [Nασσ]ιανή, ζήσασα [ἐν Θ(ε)ῷ] καλῶς καὶ ἀμέμπτως ἒτη λβ´, μῆνας ι´.
\end{otherlanguage*}

@@

The inscription says: ``Oh friend, you see here the tomb of Nassiane, Christian, ripe, sweet, fond of her husband, who
competed for virtue with Penelope. Here lies Nassiane, she lived well and blamelessly in God (?), 32 years and 10
months''. This inscription documents such as epigraphic formulas affected by religious syncretism (syncretism generally
means that a complex of phenomena and concepts derive from the meeting and fusion of different religious forms) between
Pagans and Christians, for which a Christian like Nassiane could have competed in life with Penelope, the hero
Odysseus’ wife and symbol of marital devotion. Its circular shape suggested that it was a table for the ritual of
refrigerium, the funeral feast that symbolically was consumed with the dead, even this syncretic practice \ dates back
to the pagan world \citep{Sgarlata2013}, \citep{Scandurra2014}.

@@

\noindent As discussed elsewhere (Bonacini 2013 and 2014), Google is undoubtedly the most active entity in the world committed to
preservation, dissemination and promotion of cultural heritage, well above any public institution, through an
unparalleled campaign of digitization open to users’ collaboration. This could happen because Google itself has an
incomparable capacity of economic investment. Even large international projects of digitization (such as Europeana
itself) \ are not able to compete with Google.

@@

This publication has been made on concession of the Regional Minister of Cultural Heritage and Sicilian Identity;
archaeological findings and plans presented here are exclusive property of the ``Paolo Orsi'' Regional Archaeological
Museum in Syracuse and it not allowed to be copied or duplicated by any means.

>>>>

\title{Digital enhancement of the ``Paolo Orsi'' museum: a Google Street View 360° pilot project tour}
\author[catania]{Elisa Bonacini\corref{first}}
\cortext[first]{Corresponding author. Email: e\_bonacini@hotmail.com; elisa.bonacini@unict.it}
\address[catania]{Humanities Department, University of Catania; IEMEST (Istituto Euro Mediterraneo di Scienze e Tecnologie), Palermo, Italy}
\date{2015-12-11}
\begin{document}

@@

Sicily has the highest number of UNESCO heritage sites (7/51 in total)\footnote{1997: Valley of Temples in Agrigento;
1997: Villa del Casale; 2000: Eolian Islands; 2002: Late Baroque Towns of the Val di Noto (South-Eastern Sicily); 2005:
Syracuse and the rock necropolisof Pantalica; 2013: Mount Etna; 2015: Arab-Norman sites, Palermo and the Cathedral
Churches of Cefalù and Monreale.} and of UNESCO intangible cultural heritage (3/6)\footnote{2008: Opera dei Pupi,
Sicilian puppet theatre; 2013: Mediterranean diet (transnational); 2014: Traditional agricultural practice of
cultivating the `vite ad alberello' (head-trained bush vines) of the community of Pantelleria.} in Italy and in the
world. Infact, Sicily has some of the largest and most important archaeological sites in the world: the temple of
Concordia in Agrigento, for its exceptional state of preservation has become the symbol of UNESCO itself.

Despite of this, Sicilian cultural heritage struggles to be present on Google’s platforms such as Google Art Project and
others, as it should, compared to other Italian cultural sites \citep{Bonacini2013, Bonacini2014}.

@@

In Street View Gallery,\footnote{\url{https://www.google.com/maps/views/streetview?gl=us}.} which now has contributed a great
number of users from all over the world, thousands are the spherical and geo located photos of Italian places. However,
tightening the selection to ``Landmarks of Italy'',
\footnote{\url{https://www.google.com/maps/views/streetview/italy-highlights?gl=us}.} Sicily has only 9 spherical photos,
showing the beaches of the Aeolian Islands (7), the islands of Favignana (1) and Marettimo (1).

Among Art Project’s 605 museum collections,\footnote{\url{https://www.google.com/maps/views/streetview/art-project?gl=it}.} 47
are Italian, especially relevant to Rome, Turin, Venice and Milan. One of these, not the best artistic production, is a
Sicilian contemporary collection, relevant to the International Festival of Street Artists in Giardini Naxos (Me). No
other Sicilian museum, collection or archaeological site has been included in Art Project.

Browsing on the Street View Gallery, 21 sites are all over inscribed on the World Wonders
Project:\footnote{\url{https://www.google.com/maps/views/streetview/world-wonders-project?gl=it}.} for Italy, only Pompeii
and the historic center of Florence are inscribed in this still restricted list. Many more sites can be visited
virtually from the same project linked on Google Cultural Institute’s
website:\footnote{\url{www.google.com/intl/it/culturalinstitute/worldwonders/}. } 172 sites in the world, 22 in Italy and,
finally, 2 of them in Sicily. Until last year the Unesco site of the baroque town Val di Noto was the only one in
Sicily; currently Mount Etna, has been added.

@@

Art Project, for example, allows you to view points of interest along the path in a museum, but captions are short and
photos are static. A pilot initiative has been recently launched featuring the possibility to in-browse almost 300 3D
photos of objects from the collections of six cultural institutions in the world. However 242 of these consist of scans
of animal skulls (pieces of nature, not works of art) from the California Academy of Science collection; 22 objects of
art come from Museo d’Arte Orientale (the only Italian cultural institution to join the project until now). These 3D
objects, described by short captions, are available to users to be rotated and zoomed. The aim of this new Google’s
project is to begin building the most important and largest database of 3D scans of art works worldwide.

@@

As for the displaying on Google Maps of the 360° virtual tour of the objects, we must specify that Maps so far does not
support the integration of menus, captions, photos, video, info inside the Street View virtual tour technology.
Infact, it is only allowed navigation in 360°, i.e. a virtual walk.

@@

The B2 Sector introduces the visitor to the archaeological finds from the city of Syracuse, from its foundation to
classical age. Here the most important spaces are those dedicated to the architecture of archaic and classical temples
(Athenaion, Ionic temple, Olympeion and Apollonion), to the statuary and the terracotta findings from urban excavations
during the last decades (Piazza Duomo, Ortigia, Piazza della Vittoria) and to the urban and extraurban necropolis
(Fusco, Giardino Spagna) with rich funerary kits. \

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img004.jpg}
\caption{360° tour of Sector D: the Venus Landolina.}
\label{fig:4}
\end{figure}

@@

chrismon \begin{otherlanguage*}{greek} ς χριστ\end{otherlanguage*}<\begin{otherlanguage*}{greek} ι\end{otherlanguage*}>\begin{otherlanguage*}{greek}ανῆς σεμνῆς ἀγανóφρονος [ἠδ]ὲ φιλάνδρου Nασσιανῆς τύμβον εἰσορᾷς, φίλε, κίμενον [ὣδε ἥ]τις σεμνοσύνῃσιν [ἐριζε]το Πηνελοπίῃ\end{otherlanguage*} palma chrismon \begin{otherlanguage*}{greek}ἐνθάδε κῖτε [Nασσ]ιανή, ζήσασα [ἐν Θ(ε)ῷ] καλῶς καὶ ἀμέμπτως ἒτη λβ´, μῆνας ι´.
\end{otherlanguage*}

@@

The inscription says: ``Oh friend, you see here the tomb of Nassiane, Christian, ripe, sweet, fond of her husband, who
competed for virtue with Penelope. Here lies Nassiane, she lived well and blamelessly in God (?), 32 years and 10
months''. This inscription documents such as epigraphic formulas affected by religious syncretism (syncretism generally
means that a complex of phenomena and concepts derive from the meeting and fusion of different religious forms) between
Pagans and Christians, for which a Christian like Nassiane could have competed in life with Penelope, the hero
Odysseus’ wife and symbol of marital devotion. Its circular shape suggested that it was a table for the ritual of
refrigerium, the funeral feast that symbolically was consumed with the dead, even this syncretic practice \ dates back
to the pagan world \citep{Sgarlata2013, Scandurra2014}.

@@

\noindent As discussed elsewhere \citep{Bonacini2013, Bonacini2014}, Google is undoubtedly the most active entity in the world committed to
preservation, dissemination and promotion of cultural heritage, well above any public institution, through an
unparalleled campaign of digitization open to users’ collaboration. This could happen because Google itself has an
incomparable capacity of economic investment. Even large international projects of digitization (such as Europeana
itself) \ are not able to compete with Google.

@@

This publication has been made on concession of the Regional Minister of Cultural Heritage and Sicilian Identity;
archaeological findings and plans presented here are exclusive property of the ``Paolo Orsi'' Regional Archaeological
Museum in Syracuse and are not allowed to be copied or duplicated by any means.
